Windowed Residual-Minimization Model Reduction for Dynamical Systems

### 2019-09-19 main.tex
latest

\address[a]{Sandia National Laboratories,  Livermore, CA}
\begin{abstract}
This work proposes a \methodNameLower\ (\methodAcronym) model-reduction
	formulation for dynamical systems. The proposed approach minimizes the
	time-continuous full-order-model residual over a low-dimensional 
	trial subspace across a sequence of time windows and addresses key deficiencies in existing model-reduction approaches; e.g., the time step and time scheme dependence of least-squares Petrov--Galerkin (LSPG), the exponentially growing error bounds present in Galerkin and LSPG, and the offline cost of space--time LSPG. 
We consider two types of trial subspaces
	within the proposed framework:
	one that reduces the spatial dimension of the full-order model, and one that
	reduces both the spatial and temporal dimensions of the full-order model. For both types of trial spaces, we consider two differing solution techniques: direct (i.e., discretize-then-optimize) and indirect (i.e., optimize-then-discretize). Critically, we show that various instances of this framework yield existing model-reduction approaches, namely Galerkin, Least-Squares Petrov–Galerkin, and Space–Time Least-Squares Petrov– Galerkin. Numerical experiments on the compressible Euler and Navier--Stokes equations demonstrate that, at an increased computational cost, the \methodAcronym\ formulation can yield more accurate and physically relevant solutions than the Galerkin and Least-Squares Petrov--Galerkin approaches.

@@

Residual minimization methods were extended-to and developed-for reduced-order models by a variety of researchers. Similar to LSFEM, residual minimization approaches for reduced-order models comprise finding the solution within a low-dimensional trial space that minimizes the residual norm of the full-order model. Residual minimization approaches were first applied to static (i.e., no time-dependence) reduced-order models in~\cite{legresley_1,legresley_2,legresley_3,bui_resmin_steady,rovas_thesis,carlberg_thesis,bui_thesis} and then extended to dynamical systems in~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg,carlberg_lspg_v_galerkin}. Specifically, Refs.~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg} formulated the residual minimization problem for dynamical systems by sequentially minimizing the \textit{fully discrete} residual (i.e., the residual after temporal discretization) at each time-step instance. This formulation is most commonly known as the \textit{Least--Squares Petrov--Galerkin} (LSPG) method. Ref.~\cite{carlberg_lspg_v_galerkin} provided numerous analyses on this formulation and examined connections with the standard Galerkin method and Petrov--Galerkin methods. Critically, it was shown that (1.) for various time-stepping methods LSPG has an equivalent continuous form that corresponds to a Petrov--Galerkin projection and (2.) the standard Galerkin method minimizes the time-instantaneous residual of the full-order model. Numerous numerical experiments have demonstrated that LSPG yields more stable (although not always more accurate) solutions than the Galerkin method~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg}. The common physical explanation for this improved performance is that, by minimizing the fully discrete residual, LSPG minimizes the residual over a \textit{finite-sized time-window}, rather than instantaneously as is done in the Galerkin method.

@@

The resulting approach is best viewed as a hybrid space--time method. A
variety of research has examined the formulation of space--time reduced-order
models~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,bui_thesis}
and the present approach displays commonalities with several such efforts.
First,~\cite{bui_thesis} briefly formulated a space--time least-squares ROM
and connected this formulation with optimal control; specifically the thesis
of optimize then discretize vs discretize then optimize was mentioned. This
formulation, however, was not fully developed and was eschewed for the
time-local residual minimization method (i.e., LSPG). The present work thus
formally develops (and extends) several of the concepts put forth
in~\cite{bui_thesis}. Next, Ref.~\cite{constantine_strom} developed a
space--time residual minimization formulation for model interpolation.  The
present work distinguishes itself from Ref.~\cite{constantine_strom} in that
(1.) in addition to trial spaces with space--time dimension reduction, we
consider trial spaces with spatial dimension reduction only (i.e., a classic
spatial basis is employed, opposed to a space--time basis) -- in which case
the generalized coordinates comprise a vector of functions, and (2.) we
minimize the time-continuous residual over space--time
windows;~\cite{constantine_strom} considers the solution of a fully discrete
optimization problem over the entire space--time domain. It is noted that,
similar to the current work, a time-continuous residual-minimization
formulation is used as a starting point in~\cite{constantine_strom}, and as
such thematic similarities exist between~\cite{constantine_strom} and the
present work. Lastly, Ref.~\cite{choi_stlspg} develops a space--time
formulation of the least-squares Petrov--Galerkin approach. The formulation
minimizes the \textit{fully discrete} residual over the entire space--time
domain within a given space--time trial space. 
%These space--time trial spaces are constructed through higher-order singular value decompositions (SVDs), such as the truncated high-order SVD. 
The present work distinguishes itself from Ref.~\cite{choi_stlspg} in that
(1.) we perform model reduction of the time-continuous system, (2.) we
consider trial spaces with both space--time dimension reduction and spatial
dimension reduction, and (3.) residual minimization is performed over time
windows; Ref.~\cite{choi_stlspg} considers space--time bases and minimizes the
fully-discrete residual over the entire domain.

@@

 The novel contributions and findings of the present work are:
\begin{enumerate}
%\item The development of a time-continuous residual minimization framework for reduced-order models.
\item The development of the \methodName\ (\methodAcronym) approach for dynamical systems. The approach sequentially minimizes the time-continuous full-order model residual over time windows.
\item The derivation of the stationary conditions for \methodAcronymROMs\ leveraging spatial and space--time dimension reduction trial spaces. The case of spatial dimension reduction trial spaces is of particular interest: the stationary conditions are derived via the Euler--Lagrange equations and comprise a coupled 
two-point Hamiltonian boundary value problem containing a forward and backward system. The forward system, which is forced by an auxiliary costate, evolves the generalized coordinates of the ROM. The backwards system, which is forced by the FOM residual evaluated about the ROM state, governs the dynamics of the costate. 
\item An exploration of the use of both direct (discretize then optimize) and indirect (optimize then discretize) solution techniques. Conditions under which \methodAcronym\ recovers Galerkin, LSPG, and space--time LSPG are presented.
\item Error analysis of the \methodAcronym\ approach for trial spaces with spatial dimension reduction. In particular, it is shown that the error in \methodAcronymROMs\ is bounded \textit{a priori} by a combination of the error at the start of the window and the integrated FOM ODE residual evaluated at the FOM state projected onto the trial space. 
\item Numerical experiments for trial spaces with spatial dimension reduction demonstrating two key findings:
\begin{itemize}
\item Residual minimization over larger time windows leads to more stable and physically relevant solutions as well as lower space--time residuals. 
\item Minimizing the residual over a larger time window does not necessarily lead to a more accurate solution in the $\elltwo$ norm. Residual 
minimization over an intermediary-sized time-window leads to the lowest error in the $\elltwo$ norm.
\end{itemize}
\end{enumerate}

@@

\subsection{Projection-Based Reduced-Order Models}
Projection-based ROMs generate approximate solutions to the FOM ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial space. For dynamical 
systems, two distinct types of trial spaces are used: space--time trial spaces with spatial dimension reduction (\spatialAcronym) and space--time trial spaces with space--time dimension reduction (\spaceTimeAcronym). \spatialAcronym\ trial spaces are most commonly employed and are used in, e.g., the Galerkin and Least-Squares Petrov--Galerkin (LSPG) methods. \spaceTimeAcronym\ trial spaces are typically 
employed in space--time ROMs. The following subsections outline these two trial spaces, along with the Galerkin, LSPG, and 
space--time LSPG method.

\subsection{\spatialAcronym\ Trial Spaces}
\spatialAcronym\ trial spaces approximate the FOM ODE solution, $\stateFOM$, with an approximate solution, $\approxstate$, that lives in a $K$-dimensional affine trial space, 
\begin{align}\label{eq:spatial_subspace}
& \stspaceS \defeq \trialspace \otimes \timeSpace,\\ 
& \trialspace \defeq \text{Range}(\basismat) +  \stateIntercept,  \label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, \label{eq:spatial_subspace3} 
\end{align}
where we refer to $\basismat \in \RRStar{\romdim}{\fomdim}$, with $\RRStar{\romdim}{\fomdim}$ denoting the set  $\{ \mathbf{X} \in \RR{\fomdim \times \romdim} \big| \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$, as the spatial trial basis matrix, $\trialspace$ as the spatial trial space, and $\stateIntercept \in \mathbb{R}^{\fomdim}$ as the reference state. 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$ are typically constructed through, e.g., POD~\cite{berkooz_turbulence_pod}, the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}, etc. 
ROMs leveraging \spatialAcronym\ trial spaces approximate the state as,
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) \defeq \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$, with $\genstate : [0,T] \rightarrow \mathbb{R}^{\romdim}$, are the generalized coordinates. Substituting the approximation given in~\eqref{eq:affine_trialspace} into~\eqref{eq:FOM} and performing orthogonal $\elltwo$ projection of the initial conditions onto the trial space, one obtains,
\begin{equation}\label{eq:g_truncation}
\basisspace \frac{d }{dt}\genstateArg{}{t} = \velocity(\basisspace
\genstateArg{}{t} + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Equation~\eqref{eq:g_truncation} provides an over-determined system of $N$ equations for the $K$ generalized coordinates. Dimension reduction is predominantly performed with either the \textit{Galerkin method} or the \textit{Least-Squares Petrov--Galerkin method} (LSPG). We now outline these two methods.

@@

The Galerkin method is (traditionally) obtained by the following steps: (1.) The approximation~\eqref{eq:affine_trialspace} is substituted into~\eqref{eq:FOM}, and (2.) the resulting residual is enforced to be $\stweightingMatArg{}$-orthogonal to the trial space, with $\stweightingMatArg{ } \equiv \stweightingMatOneArg{ }^T \stweightingMatOneArg{ }$ being a positive (semi-)definite weighting matrix (commonly taken to be $\mathbf{I}$). This projection process results in what is known as the Galerkin method,

@@

The FOM O$\Delta$E solution will exactly satisfy~\eqref{eq:lms}; the ROM solution will not. The LSPG method minimizes the residual of the FOM O$\Delta$E~\eqref{eq:lms} within the trial space over each time step. %We define the discrete residual for the implicit Euler method at the $n$th time-instance as,

@@

\subsection{\spaceTimeAcronym\ Trial Spaces and Space--Time ROMs}
Recent work has investigated space--time reduced-order models~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,benner_st,bui_thesis}. Space--time methods 
are advantageous in that they allow for both spatial \textit{and} temporal dimension reduction. Further, space--time 
methods are often equipped with more favorable error bounds than non-space--time methods. The approach proposed in this work displays commonalities 
with space--time methods and, as such, we briefly outline \spaceTimeAcronym\ trial spaces and space--time LSPG.

Space--time methods leverage the \spaceTimeAcronym\ trial space $\stspaceST$,
\begin{equation}\label{eq:sttrialspace_def}
 \approxstateArgnt{} \in  \stspaceST \subseteq \RR{N} \otimes \timeSpace,
\end{equation}
and the space--time trial basis \textit{functions},
\begin{align*}
\stbasisArgnt{} &: t \mapsto \stbasis(t) \\
 &: [0,T] \rightarrow  \RR{N \times \stdim} ,%\spaceTrialSpace , \\
%&  \in \RR{N \times \stdim} \otimes \timeTrialSpaceArg{},  
\end{align*}
with $\text{Range}(\stbasis) + \stateInterceptArg{} \equiv \stspaceST$ and where $\stdim$ is the dimension of the space--time generalized coordinates.
%with $\mathbb{V}_{\stdim}( \RR{N} )$ denoting the set of orthogonal $\stdim$ frames in $\RR{N}$ (e.g., an $N \times \stdim$ matrix with orthogonal column vectors).
The state is approximated by,
\begin{equation}\label{eq:stapprox1}
 \stateFOMArg{}{t} \approx \approxstateArg{}{t}  = \stbasisArg{}{t} \stgenstate + \stateInterceptArg{},
\end{equation}
%where $\stbasisArgnt{} : \RR{\stdim} \rightarrow \spaceTrialSpace \otimes \timeTrialSpaceArg{}$ with $ \stbasisArgnt{n} \in \RR{N \times \stdim} \otimes \timeTrialSpaceArg{n}$ is the space--time trial basis, 
where $\stgenstateArg{} \in \RR{\stdim}$ are the space--time generalized coordinates. Critically, the 
space--time basis matrix comprises time-dependent \textit{functions}, while the generalized coordinates are a stationary vector. This is fundamentally different than  
\spatialAcronym\ trial spaces, in where the generalized coordinates are time-dependent functions while the basis matrix is stationary.

>>>>

\address[a]{Sandia National Laboratories,  Livermore, CA}
\begin{abstract}
This work proposes a \methodNameLower\ (\methodAcronym) model-reduction
	formulation for dynamical systems. The proposed approach minimizes the
	time-continuous full-order-model residual over a low-dimensional 
	trial subspace across a sequence of time windows and addresses key deficiencies in existing model-reduction approaches; e.g., the time step and time scheme dependence of least-squares Petrov--Galerkin (LSPG), the exponentially growing error bounds present in Galerkin and LSPG, and the offline cost of space--time LSPG. 
We consider two types of trial subspaces
	within the proposed framework:
	one that reduces the spatial dimension of the full-order model, and one that
	reduces both the spatial and temporal dimensions of the full-order model. For both types of trial subspaces, we consider two differing solution techniques: direct (i.e., discretize-then-optimize) and indirect (i.e., optimize-then-discretize). Critically, we show that various instances of this framework yield existing model-reduction approaches, namely Galerkin, Least-Squares Petrov–Galerkin, and Space–Time Least-Squares Petrov– Galerkin. Numerical experiments on the compressible Euler and Navier--Stokes equations demonstrate that, at an increased computational cost, the \methodAcronym\ formulation can yield more accurate and physically relevant solutions than the Galerkin and Least-Squares Petrov--Galerkin approaches.

@@

Residual minimization methods were extended-to and developed-for reduced-order models by a variety of researchers. Similar to LSFEM, residual minimization approaches for reduced-order models comprise finding the solution within a low-dimensional trial subspace that minimizes the residual norm of the full-order model. Residual minimization approaches were first applied to static (i.e., no time-dependence) reduced-order models in~\cite{legresley_1,legresley_2,legresley_3,bui_resmin_steady,rovas_thesis,carlberg_thesis,bui_thesis} and then extended to dynamical systems in~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg,carlberg_lspg_v_galerkin}. Specifically, Refs.~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg} formulated the residual minimization problem for dynamical systems by sequentially minimizing the \textit{fully discrete} residual (i.e., the residual after temporal discretization) at each time-step instance. This formulation is most commonly known as the \textit{Least--Squares Petrov--Galerkin} (LSPG) method. Ref.~\cite{carlberg_lspg_v_galerkin} provided numerous analyses on this formulation and examined connections with the standard Galerkin method and Petrov--Galerkin methods. Critically, it was shown that (1.) for various time-stepping methods LSPG has an equivalent continuous form that corresponds to a Petrov--Galerkin projection and (2.) the standard Galerkin method minimizes the time-instantaneous residual of the full-order model. Numerous numerical experiments have demonstrated that LSPG yields more stable (although not always more accurate) solutions than the Galerkin method~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg}. The common physical explanation for this improved performance is that, by minimizing the fully discrete residual, LSPG minimizes the residual over a \textit{finite-sized time-window}, rather than instantaneously as is done in the Galerkin method.

@@

The resulting approach is best viewed as a hybrid space--time method. A
variety of research has examined the formulation of space--time reduced-order
models~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,bui_thesis}
and the present approach displays commonalities with several such efforts.
First,~\cite{bui_thesis} briefly formulated a space--time least-squares ROM
and connected this formulation with optimal control; specifically the thesis
of optimize then discretize vs discretize then optimize was mentioned. This
formulation, however, was not fully developed and was eschewed for the
time-local residual minimization method (i.e., LSPG). The present work thus
formally develops (and extends) several of the concepts put forth
in~\cite{bui_thesis}. Next, Ref.~\cite{constantine_strom} developed a
space--time residual minimization formulation for model interpolation.  The
present work distinguishes itself from Ref.~\cite{constantine_strom} in that
(1.) in addition to trial subspaces with space--time dimension reduction, we
consider trial subspaces with spatial dimension reduction only (i.e., a classic
spatial basis is employed, opposed to a space--time basis) -- in which case
the generalized coordinates comprise a vector of functions, and (2.) we
minimize the time-continuous residual over space--time
windows;~\cite{constantine_strom} considers the solution of a fully discrete
optimization problem over the entire space--time domain. It is noted that,
similar to the current work, a time-continuous residual-minimization
formulation is used as a starting point in~\cite{constantine_strom}, and as
such thematic similarities exist between~\cite{constantine_strom} and the
present work. Lastly, Ref.~\cite{choi_stlspg} develops a space--time
formulation of the least-squares Petrov--Galerkin approach. The formulation
minimizes the \textit{fully discrete} residual over the entire space--time
domain within a given space--time trial subspace. 
%These space--time trial spaces are constructed through higher-order singular value decompositions (SVDs), such as the truncated high-order SVD. 
The present work distinguishes itself from Ref.~\cite{choi_stlspg} in that
(1.) we perform model reduction of the time-continuous system, (2.) we
consider trial subspaces with both space--time dimension reduction and spatial
dimension reduction, and (3.) residual minimization is performed over time
windows; Ref.~\cite{choi_stlspg} considers space--time bases and minimizes the
fully-discrete residual over the entire domain.

@@

 The novel contributions and findings of the present work are:
\begin{enumerate}
%\item The development of a time-continuous residual minimization framework for reduced-order models.
\item The development of the \methodNameLower\ (\methodAcronym) approach for dynamical systems. The approach sequentially minimizes the time-continuous full-order model residual over time windows.
\item The derivation of the stationary conditions for \methodAcronymROMs\ leveraging spatial and space--time dimension reduction trial subspaces. The case of spatial dimension reduction trial subspaces is of particular interest: the stationary conditions are derived via the Euler--Lagrange equations and comprise a coupled 
two-point Hamiltonian boundary value problem containing a forward and backward system. The forward system, which is forced by an auxiliary costate, evolves the generalized coordinates of the ROM. The backwards system, which is forced by the FOM residual evaluated about the ROM state, governs the dynamics of the costate. 
\item An exploration of the use of both direct (discretize then optimize) and indirect (optimize then discretize) solution techniques. Conditions under which \methodAcronym\ recovers Galerkin, LSPG, and space--time LSPG are presented.
\item Error analysis of the \methodAcronym\ approach for trial subspaces with spatial dimension reduction. In particular, it is shown that the error in \methodAcronymROMs\ is bounded \textit{a priori} by a combination of the error at the start of the window and the integrated FOM ODE residual evaluated at the FOM state projected onto the trial subspace. 
\item Numerical experiments for trial subspaces with spatial dimension reduction demonstrating two key findings:
\begin{itemize}
\item Residual minimization over larger time windows leads to more stable and physically relevant solutions as well as lower space--time residuals. 
\item Minimizing the residual over a larger time window does not necessarily lead to a more accurate solution in the $\elltwo$ norm. Residual 
minimization over an intermediary-sized time-window leads to the lowest error in the $\elltwo$ norm.
\end{itemize}
\end{enumerate}

@@

\subsection{Projection-Based Reduced-Order Models}
Projection-based ROMs generate approximate solutions to the FOM ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial subspace. For dynamical 
systems, two distinct types of trial subspaces are used: space--time trial subspaces with spatial dimension reduction (\spatialAcronym) and space--time trial subspaces with space--time dimension reduction (\spaceTimeAcronym). \spatialAcronym\ trial subspaces are most commonly employed and are used in, e.g., the Galerkin and Least-Squares Petrov--Galerkin (LSPG) methods. \spaceTimeAcronym\ trial subspaces are typically 
employed in space--time ROMs. The following subsections outline these two trial subspaces, along with the Galerkin, LSPG, and 
space--time LSPG method.

\subsection{\spatialAcronym\ Trial Subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution, $\stateFOM$, with an approximate solution, $\approxstate$, that lives in a $K$-dimensional affine trial subspace, 
\begin{align}\label{eq:spatial_subspace}
& \stspaceS \defeq \trialspace \otimes \timeSpace,\\ 
& \trialspace \defeq \text{Range}(\basismat) +  \stateIntercept,  \label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, \label{eq:spatial_subspace3} 
\end{align}
where we refer to $\basismat \in \RRStar{\romdim}{\fomdim}$, with $\RRStar{\romdim}{\fomdim}$ denoting the set  $\{ \mathbf{X} \in \RR{\fomdim \times \romdim} \big| \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$, as the spatial trial basis matrix, $\trialspace$ as the spatial trial subspace, and $\stateIntercept \in \mathbb{R}^{\fomdim}$ as the reference state. 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$ are typically constructed through, e.g., POD~\cite{berkooz_turbulence_pod}, the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}, etc. 
ROMs leveraging \spatialAcronym\ trial subspaces approximate the state as,
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) \defeq \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$, with $\genstate : [0,T] \rightarrow \mathbb{R}^{\romdim}$, are the generalized coordinates. Substituting the approximation given in~\eqref{eq:affine_trialspace} into~\eqref{eq:FOM} and performing orthogonal $\elltwo$ projection of the initial conditions onto the trial subspace, one obtains,
\begin{equation}\label{eq:g_truncation}
\basisspace \frac{d }{dt}\genstateArg{}{t} = \velocity(\basisspace
\genstateArg{}{t} + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Equation~\eqref{eq:g_truncation} provides an over-determined system of $N$ equations for the $K$ generalized coordinates. Dimension reduction is predominantly performed with either the \textit{Galerkin method} or the \textit{Least-Squares Petrov--Galerkin method} (LSPG). We now outline these two methods.

@@

The Galerkin method is (traditionally) obtained by the following steps: (1.) The approximation~\eqref{eq:affine_trialspace} is substituted into~\eqref{eq:FOM}, and (2.) the resulting residual is enforced to be $\stweightingMatArg{}$-orthogonal to the trial subspace, with $\stweightingMatArg{ } \equiv \stweightingMatOneArg{ }^T \stweightingMatOneArg{ }$ being a positive (semi-)definite weighting matrix (commonly taken to be $\mathbf{I}$). This projection process results in what is known as the Galerkin method,

@@

The FOM O$\Delta$E solution will exactly satisfy~\eqref{eq:lms}; the ROM solution will not. The LSPG method minimizes the residual of the FOM O$\Delta$E~\eqref{eq:lms} within the trial subspace over each time step. %We define the discrete residual for the implicit Euler method at the $n$th time-instance as,

@@

\subsection{\spaceTimeAcronym\ Trial Spaces and Space--Time ROMs}
Recent work has investigated space--time reduced-order models~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,benner_st,bui_thesis}. Space--time methods 
are advantageous in that they allow for both spatial \textit{and} temporal dimension reduction. Further, space--time 
methods are often equipped with more favorable error bounds than non-space--time methods. The approach proposed in this work displays commonalities 
with space--time methods and, as such, we briefly outline \spaceTimeAcronym\ trial subspaces and space--time LSPG.

Space--time methods leverage the \spaceTimeAcronym\ trial subspace $\stspaceST$,
\begin{equation}\label{eq:sttrialspace_def}
 \approxstateArgnt{} \in  \stspaceST \subseteq \RR{N} \otimes \timeSpace,
\end{equation}
and the space--time trial basis \textit{functions},
\begin{align*}
\stbasisArgnt{} &: t \mapsto \stbasis(t) \\
 &: [0,T] \rightarrow  \RR{N \times \stdim} ,%\spaceTrialSpace , \\
%&  \in \RR{N \times \stdim} \otimes \timeTrialSpaceArg{},  
\end{align*}
with $\text{Range}(\stbasis) + \stateInterceptArg{} \equiv \stspaceST$ and where $\stdim$ is the dimension of the space--time generalized coordinates.
%with $\mathbb{V}_{\stdim}( \RR{N} )$ denoting the set of orthogonal $\stdim$ frames in $\RR{N}$ (e.g., an $N \times \stdim$ matrix with orthogonal column vectors).
The state is approximated by,
\begin{equation}\label{eq:stapprox1}
 \stateFOMArg{}{t} \approx \approxstateArg{}{t}  = \stbasisArg{}{t} \stgenstate + \stateInterceptArg{},
\end{equation}
%where $\stbasisArgnt{} : \RR{\stdim} \rightarrow \spaceTrialSpace \otimes \timeTrialSpaceArg{}$ with $ \stbasisArgnt{n} \in \RR{N \times \stdim} \otimes \timeTrialSpaceArg{n}$ is the space--time trial basis, 
where $\stgenstateArg{} \in \RR{\stdim}$ are the space--time generalized coordinates. Critically, the 
space--time basis matrix comprises time-dependent \textit{functions}, while the generalized coordinates are a stationary vector. This is fundamentally different than  
\spatialAcronym\ trial subspaces, in where the generalized coordinates are time-dependent functions while the basis matrix is stationary.

Substituting the approximation~\eqref{eq:stapprox1} into the FOM ODE~\eqref{eq:FOM},
\begin{equation}\label{eq:st_fom}
\frac{d }{dt} \stbasisArg{}{t} \stgenstate =  \velocity(\stbasisArg{}{t} \stgenstate + \stateIntercept,t) , \qquad \genstate(0) =\mathbb{P}(\stateFOM)(0), \qquad t \in [0,T],
\end{equation}
where $\mathbb{P} : \RR{N} \otimes \timeSpace \rightarrow \stspaceST$ is an orthogonal projection onto the trial subspace (e.g., $\elltwo$).
Equation~\eqref{eq:st_fom} is an over-determined algebraic system for the $\stdim$ unknowns. We now outline the Space--Time Least-Squares Petrov--Galerkin method~\cite{choi_stlspg}. 
The space--time Galerkin method is noted to additionally be a possibility for model reduction. As the space--time Galerkin method does not associate with any residual minimization 
principle, however, we do not outline it here.
\subsubsection{Space--Time LSPG} 
As in LSPG, Space--Time LSPG (ST-LSPG)~\cite{choi_stlspg} minimizes the fully discrete residual, with the key difference being that ST-LSPG minimizes the residual over the entire time domain. For brevity, ST-LSPG is outlined only for linear multistep methods.
%As ST-LSPG acts on the discrete solution, we first note that the discrete FOM solution lives in $\RR{N} \otimes \RR{N_t}$. ST-LSPG then seeks solutions in the trial space,
%$$\stspaceDiscrete \defeq \spaceTrialSpace \otimes \timeTrialSpaceDiscrete \subseteq \RR{N} \otimes \RR{N_t},$$
%where $\timeTrialSpaceDiscrete$ is the discrete temporal trial space. It is noted that this trial space is a specific instance of the trial space~\ref{eq:sttrialspace_def} 
%As before, we define a discretization of the continuous time-domain $[0,T]$ into $N_t$ time instances $t^n = n\Delta t, n=1,2,\ldots,N_t$ with a constant time-step $\Delta t = T/N_t$. We define now the space--time discrete residual for the implicit Euler method as,
We first define the space--time discrete residual for linear multistep methods,
\begin{align*}
\residST_{\text{LMS}} & \vcentcolon (\stateyDiscreteArg{N_t}, \ldots ,\stateyDiscreteArg{1};\stateFOMIC) \mapsto \begin{bmatrix}
%\frac{\stateyDiscreteArg{N_t} - \stateyDiscreteArg{N_t-1}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t}) \\
\residLMSArg{N_t}(\stateyDiscreteArg{N_t};\stateyDiscreteArg{N_t-1},\ldots,\stateyDiscreteArg{N_t - k^{N_t}}) \\ 
%\frac{\stateyDiscreteArg{N_t-1} - \stateFOMDiscreteArg{N_t-2}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t-1}) \\
\vdots \\
%\frac{\stateyDiscreteArg{1} - \stateFOMIC }{\Delta t} -  \velocity(\stateyDiscreteArg{1}) \\
\residLMSArg{1}(\stateyDiscreteArg{1}; \stateFOMIC) 
\end{bmatrix}  \\
& \vcentcolon \RR{\fomdim} \otimes \RR{N_t+1} \rightarrow \RR{\fomdim N_t}. 
\end{align*}
For convenience we define the same residual with the generalized coordinates as the inputs,
\begin{align*}
\genresidST_{\text{LMS}} & \vcentcolon (\stgenstatey ;\stateFOMIC) \mapsto \begin{bmatrix}
%\frac{\stateyDiscreteArg{N_t} - \stateyDiscreteArg{N_t-1}  }{\Delta t} -  \velocity(\stateyDiscreteArg{N_t}) \\
\residLMSArg{n}\bigg( \stbasisArg{}{t^{N_t}}\stgenstatey + \stateIntercept; \stbasisArg{}{t^{N_t-1}}\stgenstatey + \stateIntercept ,\ldots,  \stbasisArg{}{t^{N_t-k^{N_t}}}\stgenstatey + \stateIntercept \bigg) \\ 
\vdots \\
\residLMSArg{n}\bigg( \stbasisArg{}{t^1}\stgenstatey + \stateIntercept; \stateFOMIC \bigg) \\ 
\end{bmatrix}  \\
& \vcentcolon \RR{\stdim} \times \RR{N} \rightarrow \RR{\fomdim N_t}. 
\end{align*}
ST-LSPG is defined as,
\begin{equation*}
\stgenstate = \underset{ \stgenstatey }{\text{arg\,min}}|| \lspgWeightingSTArg{\stgenstatey}  \genresidST_{\text{LMS}}(\stgenstatey;\stateFOMIC) ||_2^2, 
\end{equation*}
where $\lspgWeightingSTArg{\cdot} \in \RR{n_{st} \times N N_t}$, with $\stdim \le n_{st} \le N N_t$, is a space--time weighting matrix.
